---
title: "predictive_analysis 2_roll_703"
author: "Priyanshu Dey 703"
date: "2026-02-05"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Problem 4 


## 4. Comparing Several Simple Linear Regressions

We analyze the `Boston` dataset to predict median home values (`medv`) using four distinct predictors: per capita crime rate (`crim`), nitrogen oxides concentration (`nox`), proportion of blacks (`black`), and lower status of the population (`lstat`).

### (a) Regression Models Output

We load the necessary libraries
```{r loading_library}
library(MASS)
```

```{r prob4_a}
data("Boston")
predictors = c("crim", "nox", "black", "lstat")
results = data.frame()

for(var in predictors){
  formula = as.formula(paste("medv ~", var))
  model = lm(formula, data = Boston)
  summ = summary(model)
  
  results = rbind(results, data.frame(
    Predictor = var,
    Intercept = round(coef(model)[1], 4),
    Coefficient = round(coef(model)[2], 4),
    Std_Error = round(summ$coefficients[2, 2], 4),
    R_Squared = round(summ$r.squared, 4),
    RSE = round(summ$sigma, 4)
  ))
}

results

```

### (b) Best Fit Analysis

Based on the table above, the model using **`lstat`** gives the best fit.

* **R-Squared**: `lstat` has the highest  of `r results[results$Predictor=="lstat", "R_Squared"]`, explaining approximately 54% of the variance in median home values.
* The next closest is `nox` at `r results[results$Predictor=="nox", "R_Squared"]` (18%).
* **Residual Standard Error (RSE)**: `lstat` also yields the lowest prediction error (RSE = `r results[results$Predictor=="lstat", "RSE"]`).

### (c) Coefficient Interpretation

We examine the coefficients for usefulness and directionality:

1. **`lstat` (Coef: `r results[results$Predictor=="lstat", "Coefficient"]`)**: This is the most useful predictor. The negative coefficient indicates that as the percentage of lower-status population increases, median home values decrease significantly.
2. **`nox` (Coef: `r results[results$Predictor=="nox", "Coefficient"]`)**: The negative relationship suggests that higher pollution (nitrogen oxides) correlates with lower home values.
3. **`crim` (Coef: `r results[results$Predictor=="crim", "Coefficient"]`)**: Crime rate shows a negative relationship with home value, though its explanatory power () is far lower than `lstat`.
4. **`black` (Coef: `r results[results$Predictor=="black", "Coefficient"]`)**: While the positive coefficient is statistically valid, this predictor has the lowest  () among the four, making it the least useful single predictor in this comparison.
