---
title: "ASSIGNMENT 6"
author: "Priyanshu Dey_703"
date: "`r Sys.Date()`"
output: word_document
  
---

## PROBLEM SHEET 6.
### Application of Subset Selection Methods

**Consider the “Hitters” dataset. We need the ISLR library to load the data. Our goal is to predict a baseball player’s salary based on the available predictors.**

```{r, message=FALSE, warning=FALSE}
library(ISLR)
library(leaps)
library(glmnet)

baseball_data = Hitters
head(baseball_data)
```

**1. Clean the data by removing any rows that have missing values.**

```{r}
clean_hitters = na.omit(baseball_data)
```

**2. Split the data randomly, keeping 80% for training and the rest for testing.**

```{r}
set.seed(123)
total_rows = nrow(clean_hitters)
train_idx = sample(1:total_rows, 210)
train_data = clean_hitters[train_idx, ]
test_data = clean_hitters[-train_idx, ]
```

**3. Perform best subset selection on the training data. We'll check models with up to 19 variables and compare them using Adjusted R-squared, BIC, and Cp.**

```{r}
subset_model = regsubsets(Salary ~ ., data = train_data, nvmax = 19)
model_summary = summary(subset_model)

best_r2 = which.max(model_summary$adjr2)
best_bic = which.min(model_summary$bic)
best_cp = which.min(model_summary$cp)

cat("Best model coefficients under BIC:\n")
print(coef(subset_model, best_bic))

cat("\nBest model coefficients under Adjusted R^2:\n")
print(coef(subset_model, best_r2))

cat("\nBest model coefficients under Cp:\n")
print(coef(subset_model, best_cp))
```

**4. Find the training MSE and test MSE for each of the best models selected above.**

```{r}

predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id=id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

# MSE for the best BIC Model
train_pred_b = predict.regsubsets(subset_model, train_data, best_bic)
test_pred_b = predict.regsubsets(subset_model, test_data, best_bic)
train_mse_bic = mean((train_data$Salary - train_pred_b)^2)
test_mse_bic = mean((test_data$Salary - test_pred_b)^2)

# MSE for the best Adjusted R^2 Model
train_pred_r = predict.regsubsets(subset_model, train_data, best_r2)
test_pred_r = predict.regsubsets(subset_model, test_data, best_r2)
train_mse_r2 = mean((train_data$Salary - train_pred_r)^2)
test_mse_r2 = mean((test_data$Salary - test_pred_r)^2)

# MSE for the best Cp Model
train_pred_c = predict.regsubsets(subset_model, train_data, best_cp)
test_pred_c = predict.regsubsets(subset_model, test_data, best_cp)
train_mse_cp = mean((train_data$Salary - train_pred_c)^2)
test_mse_cp = mean((test_data$Salary - test_pred_c)^2)

cat("For BIC model,\nTraining MSE is", train_mse_bic, "\nTest MSE is", test_mse_bic, "\n\n")
cat("For Adjusted R^2 model,\nTraining MSE is", train_mse_r2, "\nTest MSE is", test_mse_r2, "\n\n")
cat("For Cp model,\nTraining MSE is", train_mse_cp, "\nTest MSE is", test_mse_cp, "\n\n")
```

**5. Repeat the process using the forward stepwise selection method.**

```{r}
fwd_model = regsubsets(Salary ~ ., data = train_data, nvmax = 19, method = "forward")
fwd_summary = summary(fwd_model)

fwd_r2_best = which.max(fwd_summary$adjr2)
fwd_bic_best = which.min(fwd_summary$bic)
fwd_cp_best = which.min(fwd_summary$cp)

cat("Best forward model coefficients under BIC:\n")
print(coef(fwd_model, fwd_bic_best))

cat("\nBest forward model coefficients under Adjusted R^2:\n")
print(coef(fwd_model, fwd_r2_best))

cat("\nBest forward model coefficients under Cp:\n")
print(coef(fwd_model, fwd_cp_best))

# Calculating Test MSE for the best Adjusted R^2 forward model
fwd_train_preds = predict.regsubsets(fwd_model, train_data, fwd_r2_best)
fwd_test_preds = predict.regsubsets(fwd_model, test_data, fwd_r2_best)
fwd_train_mse = mean((train_data$Salary - fwd_train_preds)^2)
fwd_test_mse = mean((test_data$Salary - fwd_test_preds)^2)

cat("\nFor Forward Selection (Adjusted R^2),\nTraining MSE is", fwd_train_mse, "\nTest MSE is", fwd_test_mse, "\n")
```

**6. Repeat using the backward stepwise selection method.**

```{r}
bwd_model = regsubsets(Salary ~ ., data = train_data, nvmax = 19, method = "backward")
bwd_summary = summary(bwd_model)

bwd_r2_best = which.max(bwd_summary$adjr2)
bwd_bic_best = which.min(bwd_summary$bic)
bwd_cp_best = which.min(bwd_summary$cp)

cat("Best backward model coefficients under BIC:\n")
print(coef(bwd_model, bwd_bic_best))

cat("\nBest backward model coefficients under Adjusted R^2:\n")
print(coef(bwd_model, bwd_r2_best))

cat("\nBest backward model coefficients under Cp:\n")
print(coef(bwd_model, bwd_cp_best))

# Calculating Test MSE for the best Adjusted R^2 backward model
bwd_train_preds = predict.regsubsets(bwd_model, train_data, bwd_r2_best)
bwd_test_preds = predict.regsubsets(bwd_model, test_data, bwd_r2_best)
bwd_train_mse = mean((train_data$Salary - bwd_train_preds)^2)
bwd_test_mse = mean((test_data$Salary - bwd_test_preds)^2)

cat("\nFor Backward Selection (Adjusted R^2),\nTraining MSE is", bwd_train_mse, "\nTest MSE is", bwd_test_mse, "\n")
```

**7. Compare the results of the three methods.**

Best Subset Selection checks every possible combination to find the best model, but it takes a lot of computing power. Forward Selection adds predictors one by one, while Backward Selection starts with everything and drops predictors step by step.

Looking at the metrics, the Cp criterion usually picked very simple models with high MSE, which means they were underfitting the data. The models picked using Adjusted R-squared kept more predictors and had much better accuracy. Overall, the Backward Selection model chosen with Adjusted R-squared ended up having the lowest test MSE, making it the best for predicting the salaries. BIC models were simpler than the Adjusted R-squared ones but had slightly higher test errors.

---


## PROBLEM SHEET 7 

### Comparison of least squares, ridge regression, LASSO and Elastic Net

**We are still working with the Hitters dataset to predict Salary. First, make sure all missing values are dropped.**

```{r}
hitters_p7 = na.omit(Hitters)
```

**1. Split the dataset randomly, keeping exactly 100 observations in the test set.**

```{r}
set.seed(42) 
test_sample_idx = sample(1:nrow(hitters_p7), 100)

train_df = hitters_p7[-test_sample_idx, ]
test_df = hitters_p7[test_sample_idx, ]
```

**2. Train the model using least squares and report the training and test MSE.**

```{r}
ls_fit = lm(Salary ~ ., data = train_df)

ls_train_mse = mean(resid(ls_fit)^2)
ls_predictions = predict(ls_fit, newdata = test_df)
ls_test_mse = mean((test_df$Salary - ls_predictions)^2)

cat("Least Squares Model:\nTraining MSE:", ls_train_mse, "\nTest MSE:", ls_test_mse, "\n\n")
```

**3. Fit ridge regression models on the training set using a grid of λ values.** 
```{r}
# Setting up the matrix for glmnet
x_train = model.matrix(Salary ~ ., data = train_df)[, -1]
y_train = train_df$Salary
x_test = model.matrix(Salary ~ ., data = test_df)[, -1]

# Ridge Regression uses alpha = 0. We use cross-validation to find the best lambda.
cv_ridge = cv.glmnet(x_train, y_train, alpha = 0)
best_lambda_ridge = cv_ridge$lambda.min

ridge_fit = glmnet(x_train, y_train, alpha = 0, lambda = best_lambda_ridge)

ridge_train_preds = predict(ridge_fit, s = best_lambda_ridge, newx = x_train)
ridge_test_preds = predict(ridge_fit, s = best_lambda_ridge, newx = x_test)

ridge_train_mse = mean((y_train - ridge_train_preds)^2)
ridge_test_mse = mean((test_df$Salary - ridge_test_preds)^2)

cat("Ridge Regression Model:\nBest Lambda:", best_lambda_ridge, "\nTraining MSE:", ridge_train_mse, "\nTest MSE:", ridge_test_mse, "\n\n")
```


